TSELM: Target Speaker Extraction using Discrete Tokens and Language Models

### 2024-09-12 src/paper.tex
update

In contrast to blind speech separation, which seeks to isolate individual utterances from a mixture of known speakers, Target Speaker Extraction (TSE) focuses on extracting only the voice of the target speaker using auxiliary information. Current models are predominantly discriminative, employing masking strategies to directly minimize the distance between the estimated and clean speech signals \cite{luo2019conv,spex_plus,sepformer,sef_net}. However, these discriminative approaches often struggle to generalize to unseen data and may introduce undesirable distortions \cite{distortion}. To address these limitations, generative models aim to learn the underlying distribution of the target speech, rather than mapping mixed speech to clean speech. Recently, generative models have been shown to achieve performance comparable to that of discriminative models in blind speech separation \cite{tokensplit} and TSE \cite{target_diff}.

The discretization of audio has gained significant attention with the advancement of language models (LMs) \cite{l1, l2, l3, l4, l5}. It has achieved superior performances compared with discriminative models in
tasks such as speech enhancement \cite{selm} and blind speech separation \cite{dasb,tokensplit}.
These methods convert audio into discrete tokens, and leverages LMs to model them, thereby simplifying audio generation tasks by transforming complex regression problems into classification tasks \cite{dasb}.

Self-Supervised Learning (SSL) models such as HuBERT \cite{hubert} and WavLM \cite{wavlm} have demonstrated outstanding performances across numerous downstream tasks \cite{superb}, as they extract continuous representations rich in semantic and timbral information from speech.

>>>>

In contrast to blind speech separation, which seeks to isolate individual utterances from a mixture of known speakers, Target Speaker Extraction (TSE) focuses on extracting only the voice of the target speaker using auxiliary information. Current models are predominantly discriminative, employing masking strategies to directly minimize the distance between the estimated and clean speech signals \cite{luo2019conv,spex_plus,sepformer,sef_net}. However, these discriminative approaches often struggle to generalize to unseen data and may introduce undesirable distortions \cite{distortion}. To solve these issues, researchers have proposed generative models. This method aims to learn the underlying distribution of the target speaker's voice and use this knowledge to generate the clean speech of the target speaker from a mixture of voices rather 
than directly mapping from mixed speech to clean speech. Some generative models, like diffusion
models \cite{target_diff} and variational autoencoders (VAE) \cite{vae} have been studied. It has been demonstrated that generative models can achieve results comparable to those of discriminative models \cite{target_diff,tokensplit}.

The modeling of discretization of audio has gained significant attention with the advancement of language models (LMs) in modeling texts \cite{l1, l2, l3, l4, l5}. Researchers are trying to convert audio into discrete tokens and leverage LMs to 
model them, thereby simplifying audio generation tasks by transforming complex regression problems into classification tasks \cite{dasb}.
LMs have demonstrated its scalability and unification of different modalities in audio synthesis tasks \cite{audio_gen,deshmukh2023pengi,chen2023lauragpt}. It has also achieved superior performances compared with discriminative models in
tasks such as speech enhancement \cite{selm,mask_sr} and blind speech separation \cite{dasb,tokensplit}.

Some discretization methods \cite{selm,tokensplit,dasb} use Kmeans on the output of Self-Supervised Learning (SSL) models such as HuBERT \cite{hubert} and WavLM \cite{wavlm}. SSL models have demonstrated outstanding performances across numerous downstream tasks \cite{superb}, as they extract continuous representations rich in semantic and timbral information from speech.
